\chapter{Background and {Preliminaries}}

This chapter establishes the foundational framework for modeling and
simulating underwater vehicle-manipulator systems (UVMSs). It begins by
introducing the mathematical notation, including operators and mappings within
Lie groups and Lie algebras such as \(\mathfrak{so}(3)\) and \(\mathfrak{se}(
3)\), which are essential for describing rigid body dynamics. The concept of
reference frames is then presented, with a focus on standard frames like the
North-East-Down (NED) frame used in engineering applications. A
discussion of the special orthogonal group \(\text{SO}(3)\) follows, detailing
its role in representing rotations and its connection to the Lie algebra \(
\mathfrak{so}(3)\) through the exponential map. Building on this, the special
Euclidean group \(\text{SE}(3)\), which incorporates translations and
rotations for rigid body transformations, is introduced along with practical
tools like the \(\text{ad}\)- and \(\text{Ad}\)-operators. The chapter also
explains the Moore-Penrose pseudo-inverse and its application in redundancy
resolution and null space projections. Finally, it introduces Lagrange's
equations of motion for deriving dynamic models and highlights Runge-Kutta
methods as effective numerical techniques for simulating the dynamics and
control of UVMSs. This chapter provides the essential mathematical and
computational tools required for the analysis and design of the systems
discussed in the thesis.

% -----------------------------------------------------------------------------
\section{Notation}
\label{sec:bp:notation}

The mathematical notation is primarily based on the notation used in \cite{modsim},
\cite{fossen2021}, and \cite{sola2017}. We denote the skew-symmetric operator
of a vector $\bm{v} \in \mathbb{R}^3$ as
\begin{subequations}
\begin{align}
    [\cdot]_{\times} &: \mathbb{R}^3 \to \so \subset \mathbb{R}^{3\times 3} \\
    [\cdot]_{\times} &: \bm{v} \mapsto 
    \begin{bmatrix}
        0 & -v_3 & v_2 \\
        v_3 & 0 & -v_1 \\
        -v_2 & v_1 & 0
    \end{bmatrix},
\end{align}
\end{subequations}
where for any two vectors $\bm{u}, \bm{v} \in \mathbb{R}^3$
\begin{subequations}
\begin{align}
    \bm{u} \times \bm{v} &= [\bm{u}]_{\times} \bm{v} \\
    [\bm{u} \times \bm{v}]_{\times} &=
    [\bm{u}]_{\times} [\bm{v}]_{\times} - [\bm{v}]_{\times} [\bm{u}]_{\times}. 
\end{align}
\end{subequations}
We denote the operator that maps a twist vector to an element of the Lie algebra
$\se$ as
\begin{subequations}
\begin{align}
    [\cdot]_{\wedge} &: \mathbb{R}^6 \to \se \\
    [\cdot]_{\wedge} &: \begin{bmatrix} \bm{v} \\ \bm{w} \end{bmatrix} \mapsto
        \begin{bmatrix}
            [\bm{w}]_{\times} & \bm{v} \\
            \bm{0} & 0
        \end{bmatrix},
\end{align}
\end{subequations}
where for any two twists $\bm{\nu}, \bm{\mu} \in \mathbb{R}^6$
\begin{subequations}
\begin{align}
    \bm{\nu} \wedge \bm{\mu} &= [\bm{\nu}]_{\wedge} \bm{\mu} \\
    [\bm{\nu} \wedge \bm{\mu}]_{\wedge} &=
    [\bm{\nu}]_{\wedge} [\bm{\mu}]_{\wedge} - [\bm{\mu}]_{\wedge} [\bm{\nu}]_{\wedge}. 
\end{align}
\end{subequations}
Adopting notation from \cite{fossen2021}, the pose of a 6-DOF rigid body
is denoted by the position $\bm{p} \in \mathbb{R}^3$ and the attitude, parameterized
in Euler angles as $\bm{\Theta} \in \mathbb{R}^3$. Note that the position is
described in North-East-Down (NED) coordinates. The pose of the body is then
\begin{align}
    \bm{\eta}^T = \begin{bmatrix} \bm{p}^T & \bm{\Theta}^T \end{bmatrix}^T
\end{align}
The generalized body-fixed velocities are given by
\begin{align}
    \bm{\nu}^T = \begin{bmatrix} \bm{v}^T & \bm{\omega}^T \end{bmatrix}^T
\end{align}
where $\bm{v} \in \mathbb{R}^3$ is the linear velocity and $\bm{\omega} \in \mathbb{R}^3$
is the angular velocity.

% -----------------------------------------------------------------------------
\section{Reference Frames}
A considerable portion of the mathematical framework presented in this thesis
is built upon the concept of reference frames. As such, it is crucial to
provide a precise definition of reference frames and establish the notation
that will be utilized throughout the analysis. Although the concept of a
reference frame is relatively straightforward, a rigorous mathematical
treatment requires introducing dyadics and their associated operations. A
detailed exploration of these advanced topics is beyond the scope of this
thesis; readers interested in a comprehensive treatment are directed to \cite
{modsim} for an in-depth discussion of dyadics, vectors, and reference frames.
This chapter will instead summarize the necessary definitions and notations
used in this thesis. The notation and definitions presented in this section
are based on the conventions outlined in \cite{modsim}.

A reference frame is a coordinate system used to describe the position and
orientation of objects in space. It is defined by an origin and a set of
orthogonal unit vectors. The origin serves as a reference point for all
measurements within the frame, while the unit vectors define the axes of the
coordinate system. A vector in space, which is independent of any particular
coordinate system, can be expressed as a linear combination of the unit
vectors associated with a reference frame.  

In practical scenarios, certain standardized reference frames are frequently
used to simplify orientation and positioning tasks. For example, the North
-East-Down (NED) frame is commonly used in aerospace and marine navigation,
aligning its axes with geographical headings and altitude \cite{fossen2021}.
On the other hand, the East-North-Up (ENU) frame is often employed in
applications such as robotics and geospatial analysis, making use of easily
interpretable axes aligned with local east, north, and vertical directions.
Utilizing these conventional frames helps ensure consistency and clarity when
defining coordinate systems across various engineering and scientific domains.

For instance, consider two frames, \(a\) and \(b\), where the coordinate-free
unit vectors of frame \(a\) are denoted by  
\begin{align}
\vec{a}_i, \quad \vec{a}_j, \quad \vec{a}_k,
\end{align}  
and the unit vectors of frame \(b\) are represented as  
\begin{align}
\vec{b}_i, \quad \vec{b}_j, \quad \vec{b}_k.
\end{align}  
A vector \(\vec{v}\) can then be expressed in frame \(a\) as  
\begin{align}
\vec{v} = v^a_i \vec{a}_i + v^a_j \vec{a}_j + v^a_k \vec{a}_k, \quad v^a_l \in \mathbb{R}, \quad l \in \{i, j, k\},
\end{align}  
and similarly in frame \(b\) as  
\begin{align}
\vec{v} = v^b_i \vec{b}_i + v^b_j \vec{b}_j + v^b_k \vec{b}_k, \quad v^b_l \in \mathbb{R}, \quad l \in \{i, j, k\}.
\end{align}  

The components of \(\vec{v}\) in frames \(a\) and \(b\) are represented as column vectors \(\bm{v}^a\) and \(\bm{v}^b\), respectively:  
\begin{align}
\bm{v}^a = \begin{bmatrix} v^a_i & v^a_j & v^a_k \end{bmatrix}^T, \quad  
\bm{v}^b = \begin{bmatrix} v^b_i & v^b_j & v^b_k \end{bmatrix}^T.
\end{align}  

Transforming the column vectors \(\bm{v}^a\) and \(\bm{v}^b\) between frames
\(a\) and \(b\) is a fundamental operation in the study of reference frames.
The methodology for performing these transformations will be discussed in
the following section.  

% -----------------------------------------------------------------------------
\section{SO(3)}
\label{sec:bp:so3_se3}

The special orthogonal group in three dimensions, $\SO$, and the special
Euclidean group in three dimensions, $\SE$, are key mathematical tools in the
study of rigid body dynamics. These groups serve distinct purposes: $\SO$ is
used to represent rotations, while $\SE$ describes rigid body and spatial
transformations. This section offers a brief introduction to these groups and
highlights some of their essential properties.

We denote by $\SO$ the special orthogonal group in three dimensions, which is
the group of all $3 \times 3$ rotation matrices with determinant equal to one,
such that
\begin{align}
    \SO = \{ \bm{R} \in \mathbb{R}^{3 \times 3} \mid \bm{R}^T \bm{R} = \mathbb{I}, \det(\bm{R}) = 1 \}.
\end{align}
Intuitively, $\bm{R}^T \bm{R} = \mathbb{I}$ means that the columns of $\bm{R}$ are orthonormal,
and $\det(\bm{R}) = 1$ means that $\bm{R}$ is a proper rotation, i.e., it does not include
a reflection. The $\SO$ group is closely related to the Lie algebra $\so$
\begin{subequations}
\begin{align}
    \so &= \{ \bm{\Omega} \in \R^{3 \times 3} | \bm{\Omega}^T = -\bm{\Omega} \} \\
    [\bm{\Omega_1}, \bm{\Omega_2}]_{\so} &= \bm{\Omega_1} \bm{\Omega_2}
        - \bm{\Omega_2} \bm{\Omega_1},
\end{align}
\end{subequations}
by the exponential map
\begin{subequations}
\begin{align}
    \exp &: \so \to \SO \\
    \exp &: \bm{\Omega} \mapsto \sum_{k=0}^{\infty} \frac{1}{k!} \bm{\Omega}^k.
\end{align}
\end{subequations}
This is an important property, as it is fundamental to the definition of the
angular velocity $\bm{\omega}$ of a rigid body. Consider a rotation matrix
$\bm{R}_b^i$ that rotates a vector from the body-fixed frame $b$ to some inertial
frame $i$. The body-fixed angular velocity $\bm{\omega}_{ib}^b$ is then defined
as
\begin{align}
    \bm{\omega}_{ib}^b := \bm{R}_i^b \dot{\bm{R}}_b^i.
\end{align}
When specifying reference frames, we say that
\begin{align}
    \bm{p}^b = \bm{R}_{a}^b \bm{p}^a,
\end{align}
meaning that the rotation matrix $\bm{R}_{a}^b$ rotates a vector
$\bm{p}^a\in\R^3$ from frame $a$ to frame $b$. Some important properties
of $\bm{R}_a^b \in \SO$ are as follows \cite{modsim}:
\begin{itemize}
\item $\bm{R}_a^b = (\bm{R}_b^a)^{-1} = (\bm{R}_b^a)^T$
\item $\bm{R}_a^b \bm{R}_b^c = \bm{R}_a^c$.
\end{itemize}

Rotations in three-dimensional space can be parameterized using various methods,
including Euler angles, quaternions, angle-axis representations, and
Rodrigues parameters. In this thesis, Euler angles are used for simplicity.
Euler angles are a set of three angles that describe a sequence of three
rotations about the principal axes of a reference frame. These rotations are
performed in a specific order: typically about the $x$-axis, followed by the $y$
-axis, and finally the $z$-axis. The resulting transformation is represented by
a rotation matrix $\bm{R}$, which encodes the cumulative effect of these three
successive rotations.
Explicitly, if the three Euler angles are denoted as $\phi$, $\theta$, and $\psi$,
corresponding to rotations about the $x$-, $y$-, and $z$-axes respectively,
the rotation matrix is given by \cite{modsim}:
\begin{align}
    \bm{R} = \bm{R}_z(\psi) \bm{R}_y(\theta) \bm{R}_x(\phi),
\end{align}
where
\begin{subequations}
\begin{align}
    \bm{R}_x(\phi) &= \begin{bmatrix}
        1 & 0 & 0 \\
        0 & \cos(\phi) & -\sin(\phi) \\
        0 & \sin(\phi) & \cos(\phi)
    \end{bmatrix}, \\
    \bm{R}_y(\theta) &= \begin{bmatrix}
        \cos(\theta) & 0 & \sin(\theta) \\
        0 & 1 & 0 \\
        -\sin(\theta) & 0 & \cos(\theta)
    \end{bmatrix}, \\
    \bm{R}_z(\psi) &= \begin{bmatrix}
        \cos(\psi) & -\sin(\psi) & 0 \\
        \sin(\psi) & \cos(\psi) & 0 \\
        0 & 0 & 1
    \end{bmatrix}.
\end{align}
\end{subequations}
A disadvantage of using Euler angles is the well known problem of gimbal lock.
This occurs when two of the three axes align, causing the rotation matrix to
lose a degree of freedom. We say that the system has lost a degree of freedom
when this happens. In practice it is often possible to avoid this issue by not
controlling the system to a $\pm90^\circ$ pitch angle, or by switching to a
different angle representation when the system approaches gimbal lock. In this
work we avoid this issue by not controlling the system to these singularities.

% -----------------------------------------------------------------------------
\section{SE(3)}

The special Euclidean group in three dimensions, $\SE$, is the group of all
rigid body transformations in three-dimensional space. It is defined as
\begin{align}
    \SE := \left\{ \begin{bmatrix}
        \bm{R} & \bm{p} \\
        \bm{0} & 1
    \end{bmatrix}
        \in \R^{4 \times 4} \middle| \bm{R} \in \SO, \, \bm{p} \in \R^3
    \right\}.
\end{align}
A transformation by 
\begin{align}
    \bm{T}_b^i = \begin{bmatrix}
        \bm{R}_{b}^i & \bm{p}_{ib}^i \\
        \bm{0} & 1
    \end{bmatrix}
\end{align}
will operate on a vector $\bm{p}_{bo}^b \in \R^3$:
\begin{align}
    \begin{bmatrix}
        \bm{p}_{io}^i  \\
        1
    \end{bmatrix}
    =
    \bm{T}_b^i \begin{bmatrix}
        \bm{p}_{bo}^b \\
        1
    \end{bmatrix}
    =
    \begin{bmatrix}
        \bm{R}_{b}^i \bm{p}_{bo}^b + \bm{p}_{ib}^i \\
        1
    \end{bmatrix}.
    \label{eq:se3_transformation}
\end{align}
One can intuitively see from \autoref{eq:se3_transformation} that the transformation
$\bm{T}_b^i$
takes a vector $\bm{p}_{bo}^b$, the vector from the body-origin $b$ to the point $o$
in the $b$ frame, and transforms it to the vector $\bm{p}_{io}^i$, the vector from
the inertial-frame origin $i$ to the point $o$ described in the $i$ frame.


Consider an element $\bm{H}_{b}^i \in \SE$, which is a transformation from frame
$b$ to frame $i$. The body-fixed twist $\bm{\nu}_{ib}^b$ and the inertial-frame
twist $\bm{\nu}_{ib}^i$ are defined respectively as
\begin{subequations}
\begin{align}
    [\bm{\nu}_{ib}^b]_{\wedge} &= \left(\bm{H}_{b}^i\right)^{-1}\dot{\bm{H}}_{b}^i \label{eq:body_twist_def}\\
    [\bm{\nu}_{ib}^i]_{\wedge} &= \dot{\bm{H}}_{b}^i \left(\bm{H}_{b}^i\right)^{-1}. 
\end{align}
\end{subequations}

Similar to $\SO$, the Lie algebra $\se$ is closely related to $\SE$ by the exponential
map. The Lie algebra $\se$ is defined as
\begin{subequations}
    \label{eq:se3_algebra}
\begin{align}
    \se &= \left\{ \begin{bmatrix}
        \bm{\Omega} & \bm{v} \\
        \bm{0} & 0
    \end{bmatrix}
    \in \R^{4 \times 4}
    \middle|
    \bm{\Omega} \in \so, \, \bm{v} \in \R^3
    \right\} \\
    \left[ \Gamma_1, \Gamma_2 \right]_{\se} &= \Gamma_1 \Gamma_2 - \Gamma_2 \Gamma_1 \\
    \exp &: \se \to \SE
\end{align}
\end{subequations}
We define the matrix representation of the ad-operator
\begin{align}
    \ad_{\se} : \R^6 \to \R^{6 \times 6},
\end{align}
implicitly by requiring that for all $\bm{y} \in \R^6$
\begin{align}
    \ad_{\se}( \bm{x} ) \bm{y} = \bm{x} \wedge \bm{y}. \label{eq:ad_impl}
\end{align}
The resulting $\ad_{\se}(\bm{x})$ matrix is then
\begin{align}
    \ad_{\se} : \begin{bmatrix} \bm{v} \\ \bm{\omega} \end{bmatrix}
        \mapsto \begin{bmatrix}
            [\bm{\omega}]_{\times} & [\bm{v}]_{\times} \\
            \bm{0} & [\bm{\omega}]_{\times}
        \end{bmatrix}.
\end{align}
In a similar fashon, the matrix representation of the $\Ad$ operator
\begin{align}
    \Ad_{\SE} : \SE \to \R^{6 \times 6}
\end{align}
is defined. This operator is implicitly defined by requiring that for all
$\bm{\nu} \in R^6$, $\bm{H} \in \SE$
\begin{align}
    [\Ad(\bm{H})\bm{\nu}]_{\wedge} = \bm{H} [\bm{\nu}]_{\wedge} \bm{H}^{-1}.
\end{align}
This is essentially a mapping from body-fixed twists to inertial-frame twists,
and is extensively used when modeling UVMs. The $\Ad$-operator is then
\begin{align}
    \Ad_{\SE} : \begin{bmatrix}
        \bm{R} & \bm{p} \\
        \bm{0} & 1
    \end{bmatrix}
    \mapsto
    \begin{bmatrix}
        \bm{R} & [\bm{p}]_{\times} \bm{R} \\
        \bm{0} & \bm{R}
    \end{bmatrix}.
\end{align}
In the following chapters, the $\ad$- and $\Ad$-operators will allways refer to
the $\se$ and $\SE$ Lie algebras and groups, respectively.

% -----------------------------------------------------------------------------
\section{Pseudo-Inverse and Null Space Projections}
\label{sec:pseudoinverse}

For a given matrix $\bm{A} \in \mathbb{R}^{n\times m}$ we define the Moore-Penrose
pseudo-inverse $\bm{A}^{+} \in \mathbb{R}^{m\times n}$ as the unique matrix
satisfying the following four properties \cite{penrose1955}:
\begin{subequations}
    \begin{align}
        &\textrm{1. } & \bm{A}\bm{A}^{+}\bm{A} &= \bm{A} \\
        &\textrm{2. } & \bm{A}^{+}\bm{A}\bm{A}^{+} &= \bm{A}^{+} \\
        &\textrm{3. } & (\bm{A}\bm{A}^{+})^T &= \bm{A}\bm{A}^{+} \\
        &\textrm{4. } & (\bm{A}^{+}\bm{A})^T &= \bm{A}^{+}\bm{A}
    \end{align}
\end{subequations}
Note that in the general case, the transpose operator can be replaced by the
conjugate transpose (hermitian) operator for matrices over the complex field $\mathbb{C}$ \cite{penrose1955}.
It can be shown that for a full-column rank matrix $\bm{A}$, the pseudo-inverse is
\begin{align}
    \bm{A}^{+} = (\bm{A}^T \bm{A})^{-1} \bm{A}^T,
\end{align}
and for a full-row rank matrix $\bm{A}$, the pseudo-inverse is
\begin{align}
    \bm{A}^{+} = \bm{A}^T (\bm{A} \bm{A}^T)^{-1}.
\end{align}
In the general case, whether or not $\bm{A}$ is full rank, the pseudo-inverse can always
be computed using the singular value decomposition (SVD) of $\bm{A}$. The SVD of $\bm{A}$ is
\begin{subequations}
\begin{align}
    \bm{A} &= \bm{U} \bm{\Sigma} \bm{V}^T \\
    \bm{U} &\in \mathbb{R}^{n\times n} \textrm{ unitary} \\
    \bm{V} &\in \mathbb{R}^{m\times m} \textrm{ unitary} \\
    \bm{\Sigma} &= \operatorname{diag}(\sigma_1, \sigma_2, \ldots, \sigma_r, 0, \ldots, 0) \in \mathbb{R}^{n\times m} \\
    \sigma_1 &\geq \sigma_2 \geq \ldots \geq \sigma_r > 0 \\
    r &= \operatorname{rank}(\bm{A}).
\end{align}
\end{subequations}
Where a matrix $U$ is unitry if and only if $U^T U = U U^T = \mathbb{I}$. The pseudo-inverse
of $\bm{A}$ can then be computed as
\begin{subequations}
\begin{align}
    \bm{A}^{+} &= \bm{V} \bm{\Sigma}^{+} \bm{U}^T \\
    \bm{\Sigma}^{+} &:= \operatorname{diag}(\sigma_1^{-1}, \sigma_2^{-1}, \ldots, \sigma_r^{-1}, 0, \ldots, 0).
\end{align}
\end{subequations}
The singular value decomposition will be important when dealing with
singularity-robust task-priority control later on in this thesis.

Consider the optimization problem
\begin{align}
    \min_{\bm{x} \in \mathbb{R}^m} || \bm{A} \bm{x} - \bm{b} ||.
\end{align}
The solution to this problem is given by
\begin{align}
    \bm{x} = \bm{A}^{+} \bm{b} + (\mathbb{I} - \bm{A}^{+} \bm{A}) \bm{z},
\end{align}
where $\bm{z} \in \mathbb{R}^n$ is an arbitrary vector. The term $(\mathbb{I} - \bm{A}^{+} \bm{A}) \bm{z}$
is the null-space projection of $\bm{z}$ onto the null-space of $\bm{A}$. Parts
of this fact can bee seen by noting that
\begin{align}
    \bm{A}\left(\mathbb{I} - \bm{A}^{+} \bm{A}\right) = \bm{A} - \bm{A} = 0.
\end{align}
This is a fact that will be used extensively when formulating the task-priority
controllers in later chapters.

% -----------------------------------------------------------------------------
\section{Lagrange's Equation of Motion}
Lagrange's equation of motion is a formulation of the equations of motion of a
system in terms of generalized coordinates. The equations are derived from the
d'Alembert principle. It can be shown that the equations of motion derived from
Lagrange's equation are equivalent to the equations of motion derived from
the classical Newton-Euler equations \cite{modsim}. There are however some advantages to using
Lagrange's equation. For example, the equations are easier to derive for complex
systems where the kinetic and potential energy is known. With modern computer
algebra systems, the equations can be derived automatically, which is
a great advantage for complex systems.

Consider a system with $n$ generalized coordinates $q_1, q_2, \ldots, q_n$, and
their time derivatives $\dot{q}_1, \dot{q}_2, \ldots, \dot{q}_n$. Collect the
generalized coordinates in the vector $\bm{q} \in \mathbb{R}^n$ and the time
derivatives in the vector $\bm{\dot{q}} \in \mathbb{R}^n$. The kinetic energy
of the system is then a function $T$ of $\bm{q}$ and $\bm{\dot{q}}$
\begin{align}
    T = T(\bm{q}, \bm{\dot{q}})
\end{align}
and the potential energy is a function $U$ of $\bm{q}$
\begin{align}
    U = U(\bm{q})
\end{align}
Define the lagrangian $L$ as the difference between the kinetic and potential
energy
\begin{align}
    L(\bm{q}, \bm{\dot{q}}) = T(\bm{q}, \bm{\dot{q}}) - U(\bm{q})
\end{align}
The equations of motion are then given by \cite{modsim}:
\begin{align}
    \frac{d}{dt} \left( \nabla_{\dot{\bm{q}}} L \right) - \nabla_{\bm{q}} L = \bm{\tau}
\end{align}
where the $i$-th element of the vector $\bm{\tau}$ is the generalized force acting
on the $i$-th generalized coordinate.
\begin{align}
    \tau_i = \sum_{k=1}^N \frac{\partial \bm{r}_k}{\partial q_i} \cdot \bm{F}_k
\end{align}
where $\bm{r}_k$ is the position of the $k$-th force and $\bm{F}_k$ is the force
acting on this position. In cases where the kinetic energy is on the form
\begin{align}
    T = \frac{1}{2} \bm{\dot{q}}^T \bm{J}^T(\bm{q}) \bm{M} \bm{J}(\bm{q}) \bm{\dot{q}},
\end{align}
where $\bm{M}$ is symmetric and positive definite, the potential energy the
result of some potential field, and $\bm{J}(\bm{q})$ is the Jacobian of the
generalized coordinates, to be defined later, the equations of motion can be
written as
\begin{align}
    \bm{M}(\bm{q}) \ddot{\bm{q}} + \bm{C}(\bm{q}, \dot{\bm{q}}) \dot{\bm{q}} + \bm{g}(\bm{q}) = \bm{\tau}
\end{align}
where $\bm{M}(\bm{q})$ is symmetric and positive definite, $\bm{g}(\bm{q})$ is the
gradient of the potential field.

\section{Runge-Kutta Methods}

Runge-Kutta methods are a well-known set of numerical techniques used to solve
ordinary differential equations. They are popular because they are
straightforward to use and usually quite efficient. In this thesis, these
methods will be applied to model the dynamics of underwater vehicle-manipulator
systems, which can be described by ordinary differential equations.
When simulating controllers for these systems, it makes sense to use Runge
-Kutta methods, since they provide a convenient way to handle the required
calculations.

Consider a system of ordinary differential equations of the form
\begin{align}
    \dot{\bm{x}}(t) &= \bm{f}(\bm{x}(t), t) & \bm{x}(t_0) &= \bm{x}_0,
\end{align}
where $\bm{x}(t) : \R \to \R^n$ is the state of the system,
$\bm{f} : \R^n \times \R \to \R^n$ is a vector field, $t_0$ is the initial time,
and $\bm{x}_0$ is the initial state. In general, the solution to this system is
not known analytically, and numerical methods, such as Runge-Kutta must be used
to approximate the solution. These methods start by initializing the state to
$\bm{x}_0$ and then iteratively update the state using the vector field $\bm{f}$,
in time steps of size $h$. For explicit Runge-Kutta methods, the approximation
is given by \cite{modsim}:
\begin{subequations}
    \begin{align}
        \bm{k}_i &= \bm{f}(\bm{x}_n + h \sum_{j=1}^{i-1} a_{ij}\bm{k}_j, t_n + c_i h) &
            i &= 1, 2, \cdots, \sigma \\
        \bm{x}_{n+1} &= \bm{x}_n + h \sum_{i=1}^{\sigma} b_i \bm{k}_i,
    \end{align}
\end{subequations}
where $\{a_{ij}\}_{i,j}$, $\{b_i\}_i$, and $\{c_i\}_i$ are the coefficients of
the method, and $\sigma \in \mathbb{N}$ is the number of stages. The error $\bm{e}_{n+1}$ of
the method at time $t_{n+1}$ is given by
\begin{align}
    \bm{e}_{n+1} = \bm{x}_{n+1} - \bm{x}(t_n;t_{n+1}),
\end{align}
where $\bm{x}(t_n;t_{n+1})$ is the exact solution at time $t_{n+1}$ given the
initial condition $\bm{x}_n$ at time $t_n$. We say that the method is of order
$p$, if $p \in \mathbb{N}$ is the smallest integer such that
\begin{align}
    ||\bm{e}_{n+1}|| = \mathcal{O}(h^{p+1}),
\end{align}
where $\mathcal{O}$ is the big-O notation \cite{modsim}.
The most well-known Runge-Kutta method is the fourth-order Runge-Kutta method,
which has the following coefficients:

\begin{align}
    \begin{matrix}
        c_1 & \vline & & & & \\
        c_2 & \vline & a_{21} & & & \\
        c_3 & \vline & a_{31} & a_{32} & & \\
        c_4 & \vline & a_{41} & a_{42} & a_{43} & \\
        \hline
        & \vline & b_1 & b_2 & b_3 & b_4
    \end{matrix}
    \quad=\quad 
    \begin{matrix}
        0 & \vline & & & & \\
        \frac{1}{2} & \vline & \frac{1}{2} & & & \\
        \frac{1}{2} & \vline & 0 & \frac{1}{2} & & \\
        1 & \vline & 0 & 0 & 1 & \\
        \hline
        & \vline & \frac{1}{6} & \frac{1}{3} & \frac{1}{3} & \frac{1}{6}
    \end{matrix}
\end{align}

This method is used to simulate the dynamics of the underwater vehicle-manipulator
systems in this thesis.
